### 2020-11-21 pldi21/case-studies.tex
merge
final pldi submission

Livelits are able to address this tension.
Fig.~\ref{fig:grading}(c)\todo{sublabels} shows a Hazel program where 
the instructor alternates between programmatic and direct manipulation in several situations.

First, the instructor defines a value \li{grades} 
that records the grades for each student using a livelit, \li{\$dataframe}, 
that implements a tabular user interface. The formula bar 
allows the selected cell to be filled with an arbitrary Hazel expression. 
For the sake of demonstration, we show a cell that has been filled using another livelit, 
\li{\$slider}, in combination with symbolic manipulation.\todo{do this?}{}
The table itself displays not the expression itself but rather its value, just as in a spreadsheet.

@@

\subsection{Livelit Expansion}\label{sec:livelit-expansion}
Livelit invocations 
expand to expressions.
For example, the expansion of Fig.~\ref{fig:grading}(c) is:

@@

\subsubsection{Expansion Typing} 
\label{sec:expansion-typing}
To support abstract reasoning about the type of the expansion,
livelit definitions declare an \emph{expansion type}.
The declarations of the livelits in Fig.~\ref{fig:grading},
eliding their implementations, are:
\begin{lstlisting}[numbers=none,xleftmargin=0cm]
livelit $dataframe at Dataframe {...}
livelit $grade_cutoffs(averages: List(Float)) at 
  (.A Float, .B Float, .C Float, .D Float) {...}
livelit $slider (min: Int) (max: Int) at Int {...}
\end{lstlisting}
The expansion type of \li{\$dataframe} is \li{Dataframe},
which classifies tabular floating point data together with string row and column names (see the expansion above).
The expansion type of \li{\$grade_cutoffs} is a labeled product of grade cutoffs (field labels are written \li{.label}
rather than \li{label:} in Hazel).
Hazel displays the information in the livelit declaration when the cursor is on the livelit's name,
just as it displays typing information in other situations (not shown).\todo{cite HATRA}{}

@@

\subsubsection{Parameters}\label{sec:parameterization} 
Livelit can declare a finite number of parameters of specified types. 
For example, \li{\$grade_cutoffs} above declares one parameter,
the averages to be plotted, of type \li{List(Float)}.
Parameters are applied  
using function application notation 
as seen in Fig.~\ref{fig:grading}(c) or 
using the pipelining (i.e. reverse function application) operators, \li{<|} and \li{|>}, 
which allow multiple livelits to form dataflows (not shown).

@@

\begin{figure*}
  \begin{center}
    \begin{subfigure}[t]{0.5\textwidth}
      \centering
      \includegraphics[width=15pc]{img-filter-1.png}
      \caption{}
    \end{subfigure}\begin{subfigure}[t]{0.5\textwidth}
      \centering
      \includegraphics[width=15pc]{img-filter-2.png}
      \caption{}
    \end{subfigure}
  \end{center}
  \caption{Case Study II: Image Transformation. The image shown is determined based on the selected closure.}
  \label{fig:img-transformation}
\end{figure*}

@@

The livelit provides an expected type for each splice when it is created.
For example, the splices for the row and column keys in Fig.~\ref{fig:grading}(c)\todo{go through subfigures}{} have expected type \li{String}, 
and the remaining cells have expected type \li{Float}.
Hazel surfaces and uses the expected type when the cursor is in the splice.\todo{cite HATRA paper}{}
% If an expression of invalid type is entered, it will display in an error hole as usual,
% and in Hazel this will not prevent evaluation of other expressions (see Footnote \ref{footnote:typing}).

@@

Ensuring that clients can reason about binding while leaving expansions
 invisible 
 requires a hygiene discipline that enforces \emph{capture avoidance}
and \emph{context independence} \cite{TLMs}\todo{cite michael hygiene paper}.

@@

For example, there is one closure for \li{\$dataframe} in Fig.~\ref{fig:grading}.
It contains the value of \li{q1max}\todo{what?}{} and the other variables in scope. 
These values can be used to evaluate 
splices that use the corresponding variables, such as the cell selected in Fig.~\ref{fig:grading}(c).

@@

\subsubsection{Case Study: Live Photo Filters}\label{sec:image-transformation}
\todo{revise this once we have final figure}{}
Our next case study 
considers the situation where multiple closures are collected, and 
the workflows it enables.

@@

Motivated by this interview, 
we prototyped a collection of photo filter livelits.
One of these, \li{\$basic_adjustments}, is demonstrated in Fig.~\ref{fig:img-transformation}. 
This livelit contains two splices of type \li{Int},
one to adjust the contrast, and the other to adjust the brightness.
In this example, we have filled those splices with a \li{\$slider}, but 
as above we could enter any expression of type \li{Int}, e.g. a variable. 
The livelit shows a live preview of the transformed image.
The expansion generates calls to a browser image processing framework, 
not shown.

@@

Because the livelit appears inside a function applied (by \li{map}) twice, 
there are now two closures associated with the livelit. 
Rather than disabling live evaluation in situations like 
this, Hazel instead allows the programmer to select between the closures when 
the cursor is on the livelit expression via a simple sidebar toggle,
shown in the middle of Fig.~\ref{fig:img-transformation}.
This causes the view to change as shown. 
This makes it easy to see how the image filter being designed here will affect a
number of example images by quickly toggling between closures. 
The underlying expansion remains abstract, i.e. it refers to the image via the \li{url} variable.

>>>>

Livelits are able to address this tension.
Fig.~\ref{fig:grading} shows a Hazel program where
the instructor alternates between programmatic and direct manipulation in several situations.

First, the instructor defines a value \li{grades}
that records the grades for each student using a livelit, \li{\$dataframe},
that implements a tabular user interface. The formula bar
allows the selected cell to be filled with an arbitrary Hazel expression.
For the sake of demonstration, we show a cell that has been filled using another livelit,
\li{\$slider}, in combination with symbolic manipulation.
The table itself displays not the expression itself but rather its value, just as in a spreadsheet.

@@

\subsection{Livelit Expansion}\label{sec:livelit-expansion}
Livelit invocations
expand to expressions.
For example, the expansion of Fig.~\ref{fig:grading} is:

@@

\subsubsection{Expansion Typing}
\label{sec:expansion-typing}
To support abstract reasoning about the type of the expansion,
livelit definitions declare an \emph{expansion type}.
The declarations of the livelits in Fig.~\ref{fig:grading},
eliding their implementations, are:
\begin{lstlisting}[numbers=none,xleftmargin=0cm]
livelit $dataframe at Dataframe {...}
livelit $grade_cutoffs(averages: List(Float)) at
  (.A Float, .B Float, .C Float, .D Float) {...}
livelit $slider (min: Int) (max: Int) at Int {...}
\end{lstlisting}
The expansion type of \li{\$dataframe} is \li{Dataframe},
which classifies tabular floating point data together with string row and column names (see the expansion above).
The expansion type of \li{\$grade_cutoffs} is a labeled product of grade cutoffs (field labels are written \li{.label}
rather than \li{label:} in Hazel).
Hazel displays the information in the livelit declaration when the cursor is on the livelit's name,
just as it displays typing information in other situations (not shown) \cite{hazeltutor}.

@@

\subsubsection{Parameters}\label{sec:parameterization}
Livelit can declare a finite number of parameters of specified types.
For example, \li{\$grade_cutoffs} above declares one parameter,
the averages to be plotted, of type \li{List(Float)}.
Parameters are applied
using function application notation
as seen in Fig.~\ref{fig:grading} or
using the pipelining (i.e. reverse function application) operators, \li{<|} and \li{|>},
which allow multiple livelits to form dataflows (not shown).

@@

\begin{figure*}
  \begin{center}
    \includegraphics[width=40pc]{img-filter.png}
    % \begin{subfigure}[t]{0.5\textwidth}
    %   \centering
    %   \includegraphics[width=15pc]{img-filter-1.png}
    %   \caption{}
    % \end{subfigure}\begin{subfigure}[t]{0.5\textwidth}
    %   \centering
    %   \includegraphics[width=15pc]{img-filter-2.png}
    %   \caption{}
    % \end{subfigure}
  \end{center}
  \caption{Case Study: Image Transformation. The image shown is determined based on the selected closure.}
  \label{fig:img-transformation}
\end{figure*}

@@

The livelit provides an expected type for each splice when it is created.
For example, the splices for the row and column keys in Fig.~\ref{fig:grading} have expected type \li{String},
and the remaining cells have expected type \li{Float}.
Hazel surfaces and uses the expected type when the cursor is in the splice \cite{hazeltutor}.
% If an expression of invalid type is entered, it will display in an error hole as usual,
% and in Hazel this will not prevent evaluation of other expressions (see Footnote \ref{footnote:typing}).

@@

Ensuring that clients can reason about binding while leaving expansions
 invisible
 requires a hygiene discipline that enforces \emph{capture avoidance}
and \emph{context independence} \cite{TLMs,adamsHygiene}.

@@

For example, there is one closure for \li{\$dataframe} in Fig.~\ref{fig:grading}.
It contains the value of \li{q1\_max} and the other variables in scope.
These values can be used to evaluate
splices that use the corresponding variables, such as the cell selected in Fig.~\ref{fig:grading}.

@@

\subsubsection{Case Study: Live Image Filters}\label{sec:image-transformation}
Our next case study
considers the situation where multiple closures are collected, and
the workflows it enables.

@@

Motivated by this interview,
we prototyped a collection of photo filter livelits.
One of these, \li{\$basic_adjustments}, is demonstrated in Fig.~\ref{fig:img-transformation}. 
This livelit contains two splices of type \li{Int},
one to adjust the contrast, and the other to adjust the brightness.
In this example, we have filled those splices with \li{\$percent}, but
as above we could enter any expression of type \li{Int}, e.g. a variable.
The livelit shows a live preview of the transformed image.
The expansion generates calls to a browser image processing framework, 
not shown.

@@

Because the livelit appears inside a function applied (by \li{map}) twice, 
there are now two closures associated with the livelit. 
Hazel allows the programmer to select between the closures when 
the cursor is on the livelit expression via the sidebar toggle,
shown in the middle of Fig.~\ref{fig:img-transformation}.
This allows the client to see how the filter being designed will affect a
number of example images by quickly toggling between closures.
The underlying expansion remains abstract, i.e. it refers to the image via the \li{url} variable.

### 2020-11-21 pldi21/discussion.tex
final pldi submission

  % The strict hygiene discipline that we maintain has, we believe, substantial 
  % benefits---programmers will inevitably encounter unfamiliar livelits, and 
  % the reasoning principles that we enforce are likely to help them ``reason around''
  % the situation. However, it may be useful in certain circumstances to carefully
  % relax these restrictions. For example, some livelits might benefit from the 
  % ability to introduce bindings into splices, which is currently prevented by 
  % the capture avoidance discipline. Perhaps the 
  % editor could highlight the unusual situation.

@@

  Programming and authoring have much in common. Documents often contain structured
  information, and programs are written to manipulate structured information.
  Another future direction for livelits is as the basis for a programmable authoring 
  system, where the non-symbolic elements on the page are revealed to be code after all,
  albeit code generated by a livelit invocation that presents a more natural editing experience. 
  Taking this further, a networked collection of these
  documents could form a powerful computational wiki.
  We present this paper as a foundation for such exploration.
  %  This would 
  % require addressing the difficult problem of supporting collaborative interactions 
  % involving edits to arbitrary user interfaces. 
  % We believe that these problems are surmountable.

% \begin{itemize}
%   \item pattern matching
%   \item type splices
%   \item module splices
%   \item explicitly make bindings available in splices
%   \item better UI for closure provenance / connect with control flow / call stack better
%   \item deriving livelits from type definitions
%   \item bidirectional evaluation ala SnS
%   \item integration with structure editing more cleanly
%   \item \dots
%   \item collaborative editing?
%   \item side effects?
%   \item full screening livelits (or compositions thereof) as a way to create end-user workflows
%   \item big vision: authoring environment based on livelits
% \end{itemize}

>>>>

   The strict hygiene discipline has, we believe, substantial 
   benefits---programmers will inevitably encounter unfamiliar livelits, and 
   the reasoning principles that we enforce are likely to help them ``reason around''
   the situation. However, it may be useful in certain circumstances to 
   relax these, with the editor alerting the user to the unusual situation.

@@

  Programming and authoring have much in common. Documents often contain structured
  information, and programs are written to manipulate structured information.
  Another future direction for livelits is as the basis for a programmable authoring 
  system, where the non-symbolic elements on the page are revealed to be code after all,
  albeit code generated by a livelit invocation that presents a more natural editing experience. 
  Taking this further, a networked collection of these
  documents could form a powerful computational wiki.
  We present this paper as a foundation for such explorations.
  %  This would 
  % require addressing the difficult problem of supporting collaborative interactions 
  % involving edits to arbitrary user interfaces. 
  % We believe that these problems are surmountable.

\newpage
% \begin{itemize}
%   \item pattern matching
%   \item type splices
%   \item module splices
%   \item explicitly make bindings available in splices
%   \item better UI for closure provenance / connect with control flow / call stack better
%   \item deriving livelits from type definitions
%   \item bidirectional evaluation ala SnS
%   \item integration with structure editing more cleanly
%   \item \dots
%   \item collaborative editing?
%   \item side effects?
%   \item full screening livelits (or compositions thereof) as a way to create end-user workflows
%   \item big vision: authoring environment based on livelits
% \end{itemize}

### 2020-11-21 pldi21/implementation.tex
final pldi submission

\subsection{Text Editor Integration}
Livelits do not require the use of a structure editor. 
We have also developed a 
proof-of-concept implementation of livelits in a textual program editor, Sketch-n-Sketch \cite{sns-pldi,sns-uist},
which recently added support for the necessary mechanisms \cite{DBLP:journals/pacmpl/LubinCOC20}.
Livelit interaction causes the serialized model in the text buffer to be changed, which updates the view.
This proof-of-concept is not at feature parity with 
our main implementation, but it demonstrates that a syntax-recognizing text editor \cite{DBLP:journals/tosem/BallanceGV92,DBLP:conf/sde/HorganM84,interactive-visual-notation} 
is sufficient to support livelits, albeit with some gaps in livelit UI availability when 
there are syntax errors.

@@

\subsection{Integration into Imperative Languages}
\label{sec:imperative-langs}
Our focus in this paper was on pure languages.
Side effects pose a challenge on two fronts.
If they occur in a livelit implementation,
then the state exists at edit-time. 
The most viable approach would be to retain the model-view-update
architecture and specify that any transient mutable state will be reset between updates.
If side effects occur in livelit expansions, 
then the closure collection mechanism requires more 
care to avoid unsoundness, as described in Sec.~\ref{sec:calculus-closure-collection}.
However, these problems are surmountable, and we look forward to 
efforts to integrate livelits into imperative languages.

>>>>

\subsection{Text Editor Integration}
Livelits do not require the use of a structure editor. 
We have also developed a 
proof-of-concept implementation of livelits in a textual program editor, Sketch-n-Sketch \cite{sns-pldi,sns-uist},
which recently added support for the necessary mechanisms \cite{DBLP:journals/pacmpl/LubinCOC20}.
Livelit interaction causes the serialized model in the text buffer to be changed, which updates the view.
This proof-of-concept is not at feature parity with 
our main implementation, but it demonstrates that a syntax-recognizing text editor \cite{DBLP:journals/tosem/BallanceGV92,DBLP:conf/sde/HorganM84,interactive-visual-syntax} 
is sufficient to support livelits, albeit with some gaps in availability when 
there are syntax errors.

@@

\subsection{Integration into Imperative Languages}
\label{sec:imperative-langs}
Our focus in this paper was on pure languages.
Side effects pose a challenge on two fronts.
If they occur in a livelit implementation,
then the state exists at edit-time. 
The most viable approach would be to retain the model-view-update
architecture and specify that any transient mutable state will be reset between updates.
If side effects occur in livelit expansions, 
then the closure collection mechanism requires more 
care to avoid unsoundness, as described in Sec.~\ref{sec:calculus-closure-collection}.
% However, these problems are surmountable, and we look forward to 
% efforts to integrate livelits into imperative languages.

### 2020-11-21 pldi21/intro.tex
final pldi submission

\begin{figure*}
  \begin{minipage}[t]{0.37\textwidth}
    \begin{subfigure}[t]{\linewidth}
    \begin{snugshade}
      \vspace*{-2mm}
      \caption{\textbf{Prior Work:} Graphite \cite{Graphite}}
      \vspace*{1mm}
     \end{snugshade}
      \vspace*{-1mm}
      \includegraphics[width=\linewidth]{graphite-color-palette-green.png}
      \vspace*{-5mm}
    \end{subfigure}
    \hspace{8mm}
    \begin{subfigure}[t]{\linewidth}
     \begin{snugshade}
      \vspace*{-2mm}
      \caption{\textbf{This Paper:} Livelits are live and compositional}
      \vspace*{1mm}
     \end{snugshade}
      \vspace*{-1mm}
      \includegraphics[width=\linewidth]{slider-color-livelits.png}
    \end{subfigure}
  \end{minipage}
  \hspace{3mm}
  \begin{subfigure}[t]{0.50\textwidth}
  \begin{snugshade}
   \vspace*{-2mm}
    \caption{\textbf{Case Study}: Grading with Livelits}
    \vspace*{1mm}
     \end{snugshade}
    \vspace*{-1mm}
    \includegraphics[width=\linewidth]{grade-cutoff-livelit.png}
  \end{subfigure}
   \vspace*{-7mm}
   \caption{}
   \vspace*{-2mm}
   \label{fig:color}
\end{figure*}

Of course, we are not the first to integrate direct manipulation interfaces
into symbolic programming environments.
% Prior work on projectional editing
% and active code completion, detailed in Sec. \ref{sec:related-work},
% has also considered the problem of entering expressions
% of certain types, like \li{Color},
% using specialized GUIs integrated into a program editor.
% We detail prior work in Sec.~\ref{sec:related-work}, but 
The prior work most relevant to this paper is the {Graphite} system for Eclipse for Java,
demonstrated in Fig.~\ref{fig:color}(a) \cite{Graphite}.
Graphite allows a library provider to associate a GUI, called a \emph{palette}, with a type 
(via a Java class annotation).
Wherever an expression of this type is needed,
i.e. wherever there is a \emph{hole} of that type in the program
(as determined by Eclipse's online parser and typechecker),
the environment offers the client the option, via the code completion menu,
to interact with the palette.
Once the interaction is finished, the palette generates a
Java expression to fill the hole.
Figure~\ref{fig:color}(a), adapted from this prior work, demonstrates a color palette invoked using Graphite.
When the user presses the \li{Enter} key, the Java expression \li{new Color(173, 173, 173, 85)} is inserted at the cursor and the palette disappears.
Several related systems, such as the  
\textbf{mage} system for the Jupyter notebook environment \cite{DBLP:conf/uist/KeryRHMWP20}
and the interactive visual syntax system for Racket \cite{interactive-visual-syntax}, behave fundamentally similarly.
asdfasdfa
fasdfasdf
asdfasd\todo{placeholder for future text that might be added}

\citet{Graphite} evaluated Graphite by surveying 473 developers 
and \citet{DBLP:conf/uist/KeryRHMWP20} evaluated \textbf{mage} by interviewing 9 developers.
Both studies found that
participants viewed the proposed mechanism favorably and 
would use a suitable GUI some or all of the time.
% \footnote{When presented with a color palette,
% many participants remarked that they rarely entered colors directly into Java code,
% but rather into stylesheets.
% Other palettes, e.g. a palette
% that supported regular expression construction, were viewed as more
% suitable for Java code.
% The mechanisms being considered are suitable both for general-purpose languages
% and typed domain-specific languages like typed stylesheets, which can often be embedded into modern
% general-purpose languages.}
This and other prior work also collectively showcase a wide variety of use cases \cite{Graphite,DBLP:conf/uist/KeryRHMWP20,interactive-visual-syntax}\todo{more cites}, 
and the Graphite survey solicited dozens of additional use cases from participants,
which the authors systematically taxonomize \cite{Graphite}. 
We take these extensive empirical findings   
as evidence for, and a showcase of, 
the value of this class of mechanisms for integrating GUIs into code.

\subsection{Contributions}
\label{sec:contributions}
We turn our attention in this paper to several fundamental technical
deficiencies that limit GUI providers and clients using these prior systems.
To address these, 
we introduce a system of \emph{live literals}, or \emph{livelits}, 
demonstrated in Fig.~\ref{fig:color}(b). 
Livelits are unique in achieving all of the following properties.
(We describe which subset of these properties 
are achieved by prior systems,
including those just mentioned, 
in Sec.~\ref{sec:related-work}.)

@@

\llproperty{Hygienic Composition}
% Prior systems have limited or no support for {entering sub-expressions within the GUI}, 
% so they are useful mainly for generating expressions composed of constants,
% e.g. color constants in Fig.~\ref{fig:color}(a).
%
  Livelits support sub-expressions directly in the GUI, which we call \emph{splices} (after
   \citet{TLMs}).
  Fig.~\ref{fig:color}(b) demonstrates splicing: the RGBA components  
  are splice editors, so the client can define a variable, \li{baseline},
  to relate the color components\todo{tweak figure}{} 
  and use a slider livelit inline to specify the alpha component.
  % Similarly, Fig.~\ref{fig:grading}(b)\todo{subfigure labels}{} demonstrates a 
  %  data table livelit where each entry is a splice.

@@

\llproperty{Parameterization} Livelits can form parameterized families.
  For example, \li{\$slider} in Fig.~\ref{fig:color}(b) is parameterized by the slider's bounds.
  Parameters operate like splices, differing in that they can be partially applied in
  livelit abbreviations. For example, Fig.~\ref{fig:color}(b)\todo{add slider}{} 
  partially applies \li{\$slider} to \li{0} and \li{100} to define a \li{\$percentage} slider\todo{todo}{}.
  % \begin{lstlisting}[numbers=none]
  % let $percentage = $slider 0 100 in ...
  % \end{lstlisting}
  % Parameterization also underlies the hygiene mechanism, as we will discuss in Sec.~\ref{sec:livelit-definitions}.

@@

\llproperty{Liveness} Uniquely, livelits can evaluate splices 
  throughout the editing process 
  (i.e. in a \emph{live} manner \cite{DBLP:conf/icse/Tanimoto13}) 
  to provide feedback related to run-time behavior.
  For example, in Fig.~\ref{fig:color}(b), 
  displaying the selected color requires evaluating the RGBA
  component splices to numeric values.
  Evaluation occurs in a run-time environment (i.e. closure) determined by
  leaving the hole being filled by the livelit temporarily unfilled and then evaluating
  using a two-phased variant of the semantics for 
  live programming with typed holes developed by \citet{HazelnutLive}.
  We support live evaluation even for livelits 
  that appear inside a function. Multiple function calls lead 
  to multiple closures that the client 
   can select between.

>>>>

\begin{figure*}
  \begin{minipage}[t]{0.37\textwidth}
    \begin{subfigure}[t]{\linewidth}
    \begin{snugshade}
      \vspace*{-2mm}
      \caption{\textbf{Prior Work:} Graphite \cite{Graphite}}
       \label{fig:graphite}
      \vspace*{1mm}
     \end{snugshade}
      \vspace*{-1mm}
      \includegraphics[width=\linewidth]{graphite-color-palette-green.png}
      \vspace*{-5mm}
    \end{subfigure}
    \hspace{8mm}
    \begin{subfigure}[t]{\linewidth}
     \begin{snugshade}
      \vspace*{-2mm}
      \caption{\textbf{This Paper:} Livelits are live and compositional}
    \label{fig:color}
      \vspace*{1mm}
     \end{snugshade}
      \vspace*{-1mm}
      \includegraphics[width=\linewidth]{slider-color-livelits.png}
    \end{subfigure}
  \end{minipage}
  \hspace{10.5mm}
  \begin{subfigure}[t]{0.50\textwidth}
  \begin{snugshade}
   \vspace*{-2mm}
    \caption{\textbf{Case Study}: Grading with Livelits}
    \label{fig:grading}
    \vspace*{1mm}
     \end{snugshade}
    \vspace*{-1mm}
    \includegraphics[width=\linewidth]{grade-cutoff-livelit.png}
  \end{subfigure}
  \vspace{-4mm}
   \caption{Introductory Examples}
%   \vspace*{-2mm}
\end{figure*}

Of course, we are not the first to integrate direct manipulation interfaces
into symbolic programming environments.
% Prior work on projectional editing
% and active code completion, detailed in Sec. \ref{sec:related-work},
% has also considered the problem of entering expressions
% of certain types, like \li{Color},
% using specialized GUIs integrated into a program editor.
% We detail prior work in Sec.~\ref{sec:related-work}, but 
The prior work most relevant to this paper is the {Graphite} system for Eclipse for Java,
demonstrated in Fig.~\ref{fig:graphite} \cite{Graphite}.
Graphite allows a library provider to associate a GUI, called a \emph{palette}, with a type 
(via a Java class annotation).
Wherever an expression of this type is needed,
i.e. wherever there is a \emph{hole} of that type in the program
(as determined by Eclipse's online parser and typechecker),
the environment offers the client the option, via the code completion menu,
to interact with the palette.
Once the interaction is finished, the palette generates a
Java expression to fill the hole.
Figure~\ref{fig:graphite}, adapted from this prior work, demonstrates a color palette invoked using Graphite.
When the user presses the \li{Enter} key, the Java expression \li{new Color(173, 173, 173, 85)} is inserted at the cursor and the palette disappears.
Several related systems, such as the  
\textbf{mage} system for the Jupyter notebook environment \cite{DBLP:conf/uist/KeryRHMWP20}
behave fundamentally similarly.
Projectional editing, the visual macro system in Racket \cite{interactive-visual-syntax}, and a number of other designs
also confront this general problem of integrating GUIs and code.

\newpage
\citet{Graphite} evaluated Graphite by surveying 473 developers 
and \citet{DBLP:conf/uist/KeryRHMWP20} evaluated \textbf{mage} by interviewing 9 developers.
Both studies found that
participants viewed the proposed mechanism favorably and 
would use a suitable GUI some or all of the time.
% \footnote{When presented with a color palette,
% many participants remarked that they rarely entered colors directly into Java code,
% but rather into stylesheets.
% Other palettes, e.g. a palette
% that supported regular expression construction, were viewed as more
% suitable for Java code.
% The mechanisms being considered are suitable both for general-purpose languages
% and typed domain-specific languages like typed stylesheets, which can often be embedded into modern
% general-purpose languages.}
This and other prior work also collectively showcase a wide variety of use cases \cite{Graphite,DBLP:conf/uist/KeryRHMWP20,interactive-visual-syntax}, 
and the Graphite survey solicited dozens of additional use cases from participants,
which the authors systematically taxonomize \cite{Graphite}. 
We take these extensive empirical findings   
as evidence for, and a showcase of, 
the value of this class of mechanisms for integrating GUIs into code.
\vspace{-1mm}
\subsection{Contributions}
\label{sec:contributions}
We turn our attention in this paper to several fundamental technical
deficiencies that limit GUI providers and clients using these prior systems.
To address these, 
we introduce a system of \emph{live literals}, or \emph{livelits}, 
demonstrated in Fig.~\ref{fig:color}. 
Livelits are unique in achieving all of the following properties.
(We describe which subset of these properties 
are achieved by prior systems,
including those just mentioned, 
in Sec.~\ref{sec:related-work}.)

@@

\llproperty{Hygienic Composition}
% Prior systems have limited or no support for {entering sub-expressions within the GUI}, 
% so they are useful mainly for generating expressions composed of constants,
% e.g. color constants in Fig.~\ref{fig:color}(a).
%
  Livelits support sub-expressions directly in the GUI, which we call \emph{splices} (after
   \citet{TLMs}).
  Fig.~\ref{fig:color} demonstrates splicing: the RGBA components  
  are splice editors, so the client can define a variable, \li{baseline},
  to relate the color components 
  and use a slider livelit inline to specify the alpha component.
  % Similarly, Fig.~\ref{fig:grading}(b)\todo{subfigure labels}{} demonstrates a 
  %  data table livelit where each entry is a splice.

@@

\llproperty{Parameterization} Livelits can form parameterized families.
  For example, \li{\$slider} in Fig.~\ref{fig:color} is parameterized by the slider's bounds.
  Parameters operate like splices, differing in that they can be partially applied in
  livelit abbreviations. For example, Fig.~\ref{fig:color}
  partially applies \li{\$slider} to \li{0} and \li{100} to define a \li{\$percent} slider.
  % \begin{lstlisting}[numbers=none]
  % let $percentage = $slider 0 100 in ...
  % \end{lstlisting}
  % Parameterization also underlies the hygiene mechanism, as we will discuss in Sec.~\ref{sec:livelit-definitions}.

@@

\llproperty{Liveness} Uniquely, livelits can evaluate splices 
  throughout the editing process 
  (i.e. in a \emph{live} manner \cite{DBLP:conf/icse/Tanimoto13}) 
  to provide feedback related to run-time behavior.
  For example, in Fig.~\ref{fig:color}, 
  displaying the selected color requires evaluating the RGBA
  component splices to numeric values.
  Evaluation occurs in a run-time environment (i.e. closure) determined by
  leaving the hole being filled by the livelit temporarily unfilled and then evaluating
  using a two-phased variant of the semantics for 
  live programming with typed holes developed by \citet{HazelnutLive}.
  We support live evaluation even for livelits 
  that appear inside a function. Multiple function calls lead 
  to multiple closures that the client 
   can select between.

### 2020-11-21 pldi21/livelit-calculus.tex
final pldi submission

For simplicity, we assume that the livelit context is provided \emph{a priori} and therefore
that the expansion function is already closed and fully elaborated.
In practice, the livelit context would be controlled by a definition form in the language
that allows the definition to itself invoke livelits.
This would require a staging mechanism,
because we need to evaluate expansion functions in prior definitions to be able to
expand subsequent definitions.
There are a number of ways to support the necessary staging in practice, e.g.
via explicit staging primitives \cite{DBLP:conf/icfp/Flatt02},
by requiring that these definitions appear in separately compiled packages \cite{TLMs},
or by using live
programming mechanisms such as those available in Hazel
to evaluate ``up to'' each definition before proceeding \cite{HazelnutLive}.
% This choice is orthogonal to the mechanisms of interest in this section.

>>>>

For simplicity, we assume that the livelit context is provided \emph{a priori} and therefore
that the expansion function is already closed and fully elaborated.
In practice, the livelit context would be controlled by a definition form in the language
that allows the definition to itself invoke livelits.
This would require a staging mechanism,
because we need to evaluate expansion functions in prior definitions to be able to
expand subsequent definitions.
There are a number of ways to support the necessary staging in practice, e.g.
via explicit staging primitives \cite{DBLP:conf/icfp/Flatt02},
by requiring that these definitions appear in separately compiled packages \cite{TLMs},
or by using live
programming mechanisms such as those in Hazel
to evaluate ``up to'' each definition before proceeding \cite{HazelnutLive}.
% This choice is orthogonal to the mechanisms of interest in this section.
